Poker Chips, Earthquakes, and Computation

### 2014-10-11 finaldraft.tex
started changing stability section. rewrote first three or four paragraphs

@@

In this section we show that any finite board will reach a stable state after finitely many earthquakes.
We do so by showing that if a earthquake changes the state of a board then no sequence of earthquakes can transform the output board back into the original board. 
We will then give a bound for the maximum number of configurations that earthquakes can transform a given board into. 
Together these show that all finite boards stabilize. 
\begin{theorem}
\label{finitestability}
If a board $b$ is finite then there exists $N>0$ such that after applying a $N$ earthquakes the resultant board will be stable.
\end{theorem}

@@

\subsection{Simple Model}

A simple model of computation can be defined as follows. We assume a finite board. We specify some input nodes and some output nodes. If an input bit is to be set to 1, then we place 4 chips in the input node. We apply earthquakes to the board until it stabilizes. If an output node ever had an earthquake, then we mark that output bit as 1; otherwise, the output bit is set to 0.

A simple example is a board that computes the identity function of one output. We can compute 
\[ f(x) = x \]
By having a board with no chips, and we let the input node be the same as the output node. Then if $x = 0$, this means that the input node will not contain any chips, so the board is stable. The output node never received an earthquake, so the output is 0. If $x = 1$, then the input node gets 4 chips and an earthquake affects it. So the output node is activated, and so the output is 1.
With this simple model, there are a surprising number of things we can do. We can make wires, AND gates and OR gates (see Figure~\ref{fig:simplegates}).

\begin{figure}
\centering
\subfigure[Wire]{\includegraphics[width=0.25\linewidth]{wire.pdf}\label{fig:wire}}
\qquad\qquad
\subfigure[AND gate]{\includegraphics[width=0.25\linewidth]{andgate.pdf}\label{fig:AND}}
\qquad\qquad
\subfigure[OR gate]{\includegraphics[width=0.25\linewidth]{orgate.pdf}\label{fig:OR}}
\caption{Gates in the simple model. Green are the inputs, red are the outputs}
\label{fig:simplegates}
\end{figure}

Composing these gates, we can make circuits which represent some Boolean function. In particular we can compute any Boolean function that can be represented with a planar circuit with only AND and OR.

\begin{theorem}
There does not exists a board which computes $f(x) = \neg x$.
\end{theorem}

\begin{proof}
Suppose for contradiction that we have a board $b$ with one input $i$ and one output $o$ which computes NOT. Then we know that there exists some node which initially has over $4$ chips, since otherwise, the board would be stable and so the output never gets activated.

Let $A(0)$ be the set of nodes that get activated when $i = 0$ and $A(1)$ be the set of nodes that get activated when $i=1$. Then we claim that $A(0) \subset A(1)$. This would yield a contradiction since $o \in A(0)$ but $o \notin A(1)$. 
\end{proof}

>>>>

Generally, when we refer to an earthquake, we restrict our attention to earthquakes that actually change the state of the board.

@@

In this section, we show that any finite board will eventually become stable after a finite number of earthquakes. We first introduce a spread function that maps boards to positive numbers. We demonstrate that the spread function strictly increases after earthquakes are applied, which will imply that an earthquake irreversably changes a board configuration. In other words, we can never return to a given board arrangement.

We then give a bound on the maximum number of board configurations a finite board can ever reach. We will argue that since a finite number of boards are reachable and board states are never revisited, that eventually this process must terminate. Thus, we eventually reach a stable board where no earthquakes are allowable.

To define the spread function, we fix one coordinate on a board $b$ as the origin $(0,0)$, and we define the \emph{taxicab distance} $d(s)$ of a node $s$ on the board to be $|s_x| + |s_y|$, where $s_x$ is the $x$ coordinate of node $s$, and $s_y$ is the $y$ coordinate of node $s$. Then the spread function is
\[ \psi(b) = \sum_{\text{all nodes }s}b(s)d(s)^2 \]

The motivation for this definition is that the spread function should give a measure of how ``spread out" the chips on a board are. Intuitvely, earthquakes have the function of moving chips that are condensed in a single piile and spreading them out over a greater area. This function, which sums up the chip stack sizes weighted by the taxicab distance squared to the origin, sufficiently captures the idea of spread.

\begin{theorem}
\label{finitestability}
If a board $b$ is finite then there exists $N>0$ such that after applying a $N$ earthquakes the resultant board will be stable.
\end{theorem}

@@

\subsection{Model of Computation}
